
---
title: "schoolbuses - CQ"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

****
ASSUMPTIONS:
- To calculate percentages I'm going to only include rows for which we have arrival data, so as to make comparisons of late vs all, apples to apples.
- Furthermore, to calculate late buses, I'm only going to include those buses which were supposed to arrive before the bell. There are many buses that were supposed to arrive AFTER the bell (later sessions, etc). We'll ignore those.
```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)

data.days <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(sched_late),
    !is.na(ActualArrivalTime),
    !is.na(PlannedAnchorTime),
    !is.na(belltime)
  ) %>%
  mutate(
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1),
    plannedBeforeBell = (PlannedAnchorTime <= belltime)
  ) %>%
  mutate(
    date = dmy(date),
    five_late = ((ActualArrivalTime - belltime) + 5),
    is.bell.late = plannedBeforeBell & (bell_late > 0),
    is.five.late = plannedBeforeBell & (five_late > 0),
    is.sched.late = plannedBeforeBell & (sched_late > 0)
  )
```

****
"700 buses"
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
length(unique(rawdata$Vehicle))
length(unique(rawdata$reg_vehicle))
```

****
Percent of buses
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days %>%
  group_by(is.bell.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.five.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.sched.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()
```

****
100 minutes after the bell
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days %>%
  filter(is.bell.late) %>%
  arrange(desc(bell_late)) %>%
  head(1) %>%
  select(bell_late) %>%
  knitr::kable()
```


****
School percents
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
all.arrivals <- data.days %>%
  filter(
    school != '',
    schtype != ''
  ) %>%
  group_by(school, schtype) %>%
  summarise(arrivals = n()) %>%
  ungroup()

post.bell.arrivals <- data.days %>%
  filter(
    school != '',
    schtype != '',
    is.bell.late
  ) %>%
  group_by(school, schtype) %>%
  summarise(bell.late = n()) %>%
  ungroup()

post.bell.arrivals %>%
  inner_join(all.arrivals, by=c('school', 'schtype')) %>%
  mutate(pct = 100 * bell.late / arrivals) %>%
  arrange(desc(pct)) %>%
  filter(schtype == 'BPS') %>%
  head(20) %>%
  knitr::kable()
```

****
Number of unique routes
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
length(unique(rawdata$Route))
```


****
Most common planned time
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days %>%
  filter(plannedBeforeBell) %>%
  mutate(window = belltime - PlannedAnchorTime) %>%
  ggplot(aes(window)) +
  geom_histogram()
```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


****
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}



```


