---
title: "schoolbuses - main"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
```

****
Why is BPS using **bell_late** to determine tardiness? Why not 10 minutes before school starts? That would seem to give kids some time to get off the bus and on to their first class. Let's create a new variable, **ten_late**, and calculate overall tardiness by three definitions of late:

- **bell late**: did the bus make the school bell?
- **ten late**: did the bus arrive ten minutes before the school bell?
- **scheduled late**: did the bus arrive past its scheduled (by the software) date?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(sched_late),
    !is.na(belltime),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    date = floor_date(dmy(date), 'day'),
    ten_late = ((hm(ActualArrivalTime) - hm(belltime))/minutes(1) + 10),
    is.bell.late = bell_late > 0,
    is.ten.late = ten_late > 0,
    is.sched.late = sched_late > 0
  )

data.days %>%
  group_by(is.bell.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.ten.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.sched.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()
```

****
Let's plot weekly tardiness by these three variables.
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
bell.late <- data.days %>%
  group_by(date, is.bell.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(bell.late = n/sum(n)) %>%
  filter(is.bell.late) %>%
  select(-is.bell.late, -n)

ten.late <- data.days %>%
  group_by(date, is.ten.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(ten.late = n/sum(n)) %>%
  filter(is.ten.late) %>%
  select(-is.ten.late, -n)

sched.late <- data.days %>%
  group_by(date, is.sched.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(sched.late = n/sum(n)) %>%
  filter(is.sched.late) %>%
  select(-is.sched.late, -n)

late <- bell.late %>%
  inner_join(ten.late, by=c('date')) %>%
  inner_join(sched.late, by=c('date')) %>%
  melt(c('date'))

ggplot(late, aes(date, value, fill=variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('Percent of tardy buses by various definitions of "late"') +
  scale_y_continuous(labels=percent) +
  theme(axis.title.y = element_blank()) +
  xlab('day')
```

****
What's the average time between scheduled arrival and school start?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(belltime),
    !is.na(PlannedAnchorTime)
  ) %>%
  mutate(delta = ((hm(belltime) - hm(PlannedAnchorTime))/minutes(1)))

ggplot(data) +
  geom_histogram(aes(delta), binwidth=1) +
  scale_y_sqrt() +
  ggtitle('Time between scheduled arrival and bell') +
  xlab('minutes') +
  ylab('number of trips (square root scale)')
```

This shows that the vast majority of trips are scheduled to arrive 15 minutes before the bell. Some are scheduled to arrive after the bell, but it's such a small number - `r I(100 * nrow(data %>% filter(delta < 0))/nrow(data))`% - that it doesn't affect our analysis. 

****
How many trips arrived on time the week of October 8 2013, the day when bus drivers went on strike?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
late %>%
  filter(
    year(date) == 2013,
    month(date) == 10,
    day(floor_date(date, 'week')) == 6
  ) %>%
  ggplot(aes(date, value, fill=variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('Percent of tardy buses by various definitions of "late" for the week of October 6 2013') +
  scale_y_continuous(labels=percent) +
  theme(axis.title.y = element_blank()) +
  xlab('day')
```








