---
title: "schoolbuses - main"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
```

### Summary

In the 2013-14 school year, Boston Public School buses arrived after the bell 7% of the time. However, they arrived with less than 10 minutes before the bell, 29% of the time. What impact does this have on kids?

- If they eat breakfast before bell, then only arriving at bell is not enough.
- If they eat breakfast after bell, then it seems schools are accepting buses will be tardy, and are incorporating chronic bus lateness into the school structure. 
- According to BPS media relations director Denise Snyder, schools offer a "Grab and go" alternative for students that missed breakfast. What is in this "Grab and go" and how does it differ from regular breakfast?

### Data analysis

****
Why is BPS using **bell_late** to determine tardiness? Why not 10 minutes before school starts? That would seem to give kids some time to get off the bus and on to their first class. Let's create a new variable, **ten_late**, and calculate overall tardiness by three definitions of late:

- **bell late**: did the bus make the school bell?
- **ten late**: did the bus arrive ten minutes before the school bell?
- **scheduled late**: did the bus arrive past its scheduled (by the software) date?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(sched_late),
    !is.na(belltime),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    date = floor_date(dmy(date), 'day'),
    ten_late = ((hm(ActualArrivalTime) - hm(belltime))/minutes(1) + 10),
    is.bell.late = bell_late > 0,
    is.ten.late = ten_late > 0,
    is.sched.late = sched_late > 0
  )

data.days %>%
  group_by(is.bell.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.ten.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.sched.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()
```

****
Let's plot weekly tardiness by these three variables.
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
bell.late <- data.days %>%
  group_by(date, is.bell.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(bell.late = n/sum(n)) %>%
  filter(is.bell.late) %>%
  select(-is.bell.late, -n)

ten.late <- data.days %>%
  group_by(date, is.ten.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(ten.late = n/sum(n)) %>%
  filter(is.ten.late) %>%
  select(-is.ten.late, -n)

sched.late <- data.days %>%
  group_by(date, is.sched.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(sched.late = n/sum(n)) %>%
  filter(is.sched.late) %>%
  select(-is.sched.late, -n)

late <- bell.late %>%
  inner_join(ten.late, by=c('date')) %>%
  inner_join(sched.late, by=c('date')) %>%
  melt(c('date'))

ggplot(late, aes(date, value, fill=variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('Percent of tardy buses by various definitions of "late"') +
  scale_y_continuous(labels=percent) +
  theme(axis.title.y = element_blank()) +
  xlab('day')
```

****
Let's look at vehicles. First question: what's the distribution of bell_late by yard?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late)
  ) %>%
  mutate(is.bell.late = bell_late > 0) %>%
  group_by(yard, is.bell.late) %>%
  tally()

all_arrivals <- data %>%
  group_by(yard) %>%
  tally %>%
  rename(all = n)

data %>%
  filter(is.bell.late) %>%
  select(-is.bell.late) %>%
  inner_join(all_arrivals, by=c('yard')) %>%
  rename(late = n) %>%
  mutate(percent = 100*late/all) %>%
  ungroup() %>%
  arrange(desc(percent)) %>%
  knitr::kable()

rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    bell_late > 0
  ) %>%
  ggplot(aes(bell_late)) +
  geom_histogram() +
  scale_x_sqrt() +
  scale_y_sqrt() +
  facet_grid(yard ~ .) +
  ggtitle('Distribution of non-zero bell late by yard') +
  ylab('number of trips (square root scale)') +
  xlab('minutes after bell (square root scale)')
```

****
How many times does Vehicle != reg_vehicle?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
# calculate planned vehicle trips per year
planned_trips <- rawdata %>%
  filter(
    errortype == '',
    reg_vehicle != ''
  ) %>%
  group_by(reg_vehicle) %>%
  tally() %>%
  rename(planned = n)

unplanned_trips <- rawdata %>%
  filter(
    errortype == '',
    reg_vehicle != '',
    as.character(Vehicle) != as.character(reg_vehicle)
  ) %>%
  group_by(reg_vehicle) %>%  
  tally() %>%
  rename(unplanned = n) %>%
  ungroup() %>%
  arrange(desc(unplanned))

unplanned_trips %>%
  inner_join(planned_trips, by=c('reg_vehicle')) %>%
  mutate(percent = unplanned/planned) %>%
  arrange(desc(unplanned))
```

****
Looks like vehicle HB211 didn't make its intended trip 351 out of 366 times, or 96%! What's going on? Where is it going?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    errortype == '',
    reg_vehicle == 'HB211'
  ) %>%
  group_by(school) %>%
  tally() %>%
  knitr::kable()
```

****
Let's plot all HB211 arrivals.
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    errortype == '',
    school == 'Edison K-8',
    !is.na(PlannedStartTime),
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/hours(1),
    ActualArrivalTime = hm(ActualArrivalTime)/hours(1)
  ) %>%
  filter(ActualArrivalTime > 8) %>%
  select(date, PlannedAnchorTime, ActualArrivalTime) %>%
  melt(id=c('date')) %>%
  ggplot(aes(date, value, color=variable)) +
  geom_point(size=1)
```





