---
title: "schoolbuses"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
```

****
#### Before we begin: our data has `r I(formatC(nrow(rawdata), format='d', big.mark=','))` rows and `r I(ncol(rawdata))` columns.

****
#### Let's plot all univariate distributions.
```{r, echo=FALSE, results='asis', fig.width=10, fig.height=60, warning=FALSE, message=FALSE}

p1  <- ggplot(rawdata, aes(Route)) + geom_histogram()

d2 <- rawdata %>% group_by(Vehicle) %>% tally()
d2$Vehicle <- factor(d2$Vehicle, levels = d2$Vehicle[order(d2$n)])
p2 <- ggplot(d2, aes(Vehicle, n)) + geom_bar(stat='identity')

p3  <- ggplot(rawdata, aes(Driver)) + geom_histogram()

d4 <- rawdata %>% group_by(AnchorLocation) %>% tally()
d4$AnchorLocation <- factor(d4$AnchorLocation, levels = d4$AnchorLocation[order(d4$n)])
p4 <- ggplot(d4, aes(AnchorLocation, n)) + geom_bar(stat='identity')

p5	<- ggplot(rawdata) + geom_histogram(aes(VersatransDelay))
p6 <- rawdata %>% mutate(date = dmy(as.character(date))) %>% ggplot(aes(date)) + geom_histogram()
p7 <- rawdata %>% mutate(PlannedStartTime = hm(PlannedStartTime)/hours(1)) %>% ggplot(aes(PlannedStartTime)) + geom_histogram()
p8 <- rawdata %>% mutate(PlannedAnchorTime = hm(PlannedAnchorTime)/hours(1)) %>% ggplot(aes(PlannedAnchorTime)) + geom_histogram()
p9 <- rawdata %>% mutate(ActualArrivalTime = hm(ActualArrivalTime)/hours(1)) %>% ggplot(aes(ActualArrivalTime)) + geom_histogram()

p10  <- ggplot(rawdata) + geom_histogram(aes(students))

d11 <- rawdata %>% group_by(reg_vehicle) %>% tally()
d11$reg_vehicle <- factor(d11$reg_vehicle, levels = d11$reg_vehicle[order(d11$n)])
p11 <- ggplot(d11, aes(reg_vehicle, n)) + geom_bar(stat='identity')

d12 <- rawdata %>% group_by(yard) %>% tally()
d12$yard <- factor(d12$yard, levels = d12$yard[order(d12$n)])
p12 <- ggplot(d12, aes(yard, n)) + geom_bar(stat='identity')

p13  <- ggplot(rawdata) + geom_histogram(aes(week))

d14 <- rawdata %>% group_by(dow) %>% tally()
d14$dow <- factor(d14$dow, levels = d14$dow[order(d14$n)])
p14 <- ggplot(d14, aes(dow, n)) + geom_bar(stat='identity')

p15  <- ggplot(rawdata) + geom_histogram(aes(dateplus60))
p16	<- ggplot(rawdata) + geom_histogram(aes(run))

p17	<- ggplot(rawdata) + geom_histogram(aes(first_run))
p18	<- ggplot(rawdata) + geom_histogram(aes(sch_route))
p19	<- ggplot(rawdata) + geom_histogram(aes(sch))
p20	<- ggplot(rawdata) + geom_histogram(aes(startnumber))

d22	<- rawdata %>% group_by(type) %>% tally()
d22$type <- factor(d22$type, levels = d22$type[order(d22$n)])
p22 <- ggplot(d22, aes(type, n)) + geom_bar(stat='identity')

d23	<- rawdata %>% group_by(bps_school) %>% tally()
d23$bps_school <- factor(d23$bps_school, levels = d23$bps_school[order(d23$n)])
p23 <- ggplot(d23, aes(bps_school, n)) + geom_bar(stat='identity')

d24	<- rawdata %>% group_by(ntwk) %>% tally()
d24$ntwk <- factor(d24$ntwk, levels = d24$ntwk[order(d24$n)])
p24 <- ggplot(d24, aes(ntwk, n)) + geom_bar(stat='identity')

d25	<- rawdata %>% group_by(officer) %>% tally()
d25$officer <- factor(d25$officer, levels = d25$officer[order(d25$n)])
p25 <- ggplot(d25, aes(officer, n)) + geom_bar(stat='identity')

d26	<- rawdata %>% group_by(schtype) %>% tally()
d26$schtype <- factor(d26$schtype, levels = d26$schtype[order(d26$n)])
p26 <- ggplot(d26, aes(schtype, n)) + geom_bar(stat='identity')

d27	<- rawdata %>% group_by(school) %>% tally()
d27$school <- factor(d27$school, levels = d27$school[order(d27$n)])
p27 <- ggplot(d27, aes(school, n)) + geom_bar(stat='identity')

p28 <- rawdata %>% mutate(pmbell = hm(pmbell)/hours(1)) %>% ggplot(aes(pmbell)) + geom_histogram()
p29 <- rawdata %>% mutate(belltime = hm(belltime)/hours(1)) %>% ggplot(aes(belltime)) + geom_histogram()
p30 <- rawdata %>% mutate(amanchortime = hm(amanchortime)/hours(1)) %>% ggplot(aes(amanchortime)) + geom_histogram()

d31	<- rawdata %>% group_by(errortype) %>% tally()
d31$errortype <- factor(d31$errortype, levels = d31$errortype[order(d31$n)])
p31 <- ggplot(d31, aes(errortype, n)) + geom_bar(stat='identity')

p32 <- ggplot(rawdata) + geom_histogram(aes(tmp_late))
p33 <- ggplot(rawdata) + geom_histogram(aes(LD_late))
p34 <- ggplot(rawdata) + geom_histogram(aes(bell_late))
p35 <- ggplot(rawdata) + geom_histogram(aes(sched_late))
p36 <- ggplot(rawdata) + geom_histogram(aes(B_AllBellOntime))
p37 <- ggplot(rawdata) + geom_histogram(aes(B_AllBellLate))
p38 <- ggplot(rawdata) + geom_histogram(aes(B_bell_0_5))
p39 <- ggplot(rawdata) + geom_histogram(aes(B_bell_6_10))
p40 <- ggplot(rawdata) + geom_histogram(aes(B_bell_11_29))
p41 <- ggplot(rawdata) + geom_histogram(aes(B_bell_29))
p42 <- ggplot(rawdata) + geom_histogram(aes(B_Ontime))
p43 <- ggplot(rawdata) + geom_histogram(aes(B_LDLate))
p44 <- ggplot(rawdata) + geom_histogram(aes(B_LDLateExtreme))
p45 <- ggplot(rawdata) + geom_histogram(aes(B_sched_ontime))
p46 <- ggplot(rawdata) + geom_histogram(aes(B_sched_late))
p47 <- ggplot(rawdata) + geom_histogram(aes(B_sched_0_5))
p48 <- ggplot(rawdata) + geom_histogram(aes(B_sched_within_5))
p49 <- ggplot(rawdata) + geom_histogram(aes(B_NoArrival))
p50 <- ggplot(rawdata) + geom_histogram(aes(B_AllBellOntime_students))
p51 <- ggplot(rawdata) + geom_histogram(aes(B_bell_0_5_students))
p52 <- ggplot(rawdata) + geom_histogram(aes(B_bell_6_10_students))
p53 <- ggplot(rawdata) + geom_histogram(aes(B_bell_11_29_students))
p54 <- ggplot(rawdata) + geom_histogram(aes(B_bell_29_students))
p55 <- ggplot(rawdata) + geom_histogram(aes(B_AllBellLate_students))
p56 <- ggplot(rawdata) + geom_histogram(aes(B_Ontime_students))
p57 <- ggplot(rawdata) + geom_histogram(aes(B_LDLate_students))
p58 <- ggplot(rawdata) + geom_histogram(aes(B_LDLateExtreme_students))
p59 <- ggplot(rawdata) + geom_histogram(aes(B_NoArrival_students))
p60 <- ggplot(rawdata) + geom_histogram(aes(B_sched_ontime_students))

multiplot(
  p1, p2, p3, p4, p5, p6, p7, p8, p9, p10,
  p11, p12, p13, p14, p15, p16, p17, p18, p19, p20,
  p22, p23, p24, p25, p26, p27, p28, p29, p30,
  p31, p32, p33, p34, p35, p36, p37, p38, p39, p40,
  p41, p42, p43, p44, p45, p46, p47, p48, p49, p50,
  p51, p52, p53, p54, p55, p56, p57, p58, p59, p60,
  cols=2)
```

****
#### Are any routes consistently late? Let's take a closer look at **bell_late** and **sched_late**. Definitions: **bell_late** is the number of minutes a bus arrived after the bell. **sched_late** is the number of minutes a bus arrived after the scheduled arrival time.
```{r, echo=FALSE, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
multiplot(p34, p35, cols=2)
```

****
#### The **bell_late** histogram appears to indicate some routes are showing up hours after school starts. But maybe this is the intended behavior? What is the correlation between **bell_late** and **sched_late** for buses that arrive at least 30 minutes after school starts?
```{r, echo=FALSE, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    !is.na(sched_late),
    !is.na(bell_late)
  ) %>%
  ggplot(aes(sched_late, bell_late)) +
  geom_point()
```

****
#### See that vertical line at middle left? That tells us that there are several routes scheduled to arrive hours after **bell_late**. In other words, just because a route arrived hours after school starts doesn't make it late. There are routes that are scheduled to arrive in the afternoon. Let's focus on **sched_late** instead. Let's look at the distribution again, and we'll use a log axis on the y scale. The red line is the median.
```{r, echo=FALSE, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    !is.na(sched_late),
    !is.na(bell_late)
  )

ggplot(data) +
  geom_histogram(aes(sched_late)) +
  scale_y_log10(labels=comma) +
  geom_vline(aes(xintercept=median(sched_late, na.rm=T)), color='red', linetype='dashed', size=1)
```

****
#### This tells us that on average, buses arrive at their destination one minute after scheduled arrival. But the distribution has a long tail - thousands of buses are arriving over 30 minutes late. Which ones?
```{r, echo=FALSE, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}



```







