---
title: "schoolbuses"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
lunches <- read.csv('input/school_enrollment_by_special_population.csv', strip.white=TRUE, stringsAsFactors=FALSE, row.names=NULL)
```

****
Before we begin: our data has `r I(formatC(nrow(rawdata), format='d', big.mark=','))` rows and `r I(ncol(rawdata))` columns.

****
## Global assumptions: we'll call routes that run at least 150 trips per year, **year-round**.

****
Let's start by looking at **bell_late** and **sched_late**. Definitions: **bell_late** is the number of minutes a trip arrived after the bell. **sched_late** is the number of minutes a trip arrived after the scheduled arrival time.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  select(Route, date, bell_late, sched_late) %>%
  melt(id = c('Route', 'date')) %>%
  ggplot() +
  geom_histogram(aes(value)) +
  facet_grid(. ~ variable) +
  scale_y_sqrt(label=comma) +
  ggtitle('Distribution of trip lateness') +
  xlab('minutes') +
  ylab('number of trips (square root scale)')
```

****
The **bell_late** histogram appears to indicate some trips are showing up hours after school starts. But maybe this is the intended behavior?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE, cache=TRUE}
rawdata %>%
  filter(
    !is.na(sched_late),
    !is.na(bell_late)
  ) %>%
  ggplot(aes(sched_late, bell_late)) +
  geom_point() +
  ggtitle('Sched_late by bell_late') +
  ylab('bell_late (minutes)') +
  xlab('sched_late (minutes)')
```

****
See that vertical line at middle left? That tells us that there are several trips scheduled to arrive hours after **bell_late**. In other words, just because a trip arrived hours after school starts doesn't make it late. There are trips that are scheduled to arrive in the afternoon. But maybe most trips arrive in the morning? What's the average number of trips per route?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
tripsPerRoute <- rawdata %>%
  group_by(Route) %>%
  summarise(yearly.trips = n())

ggplot(tripsPerRoute) +
  geom_histogram(aes(yearly.trips), binwidth=1) +
  scale_y_sqrt() +
  geom_vline(aes(xintercept=median(yearly.trips, na.rm=T)), color='red', linetype='dashed', size=1) +
  ggtitle(str_c('Distribution of yearly trips per route. Median is ', median(tripsPerRoute$yearly.trips) ,' trips.')) +
  xlab('yearly trips') +
  ylab('number of routes (square root scale)')
```

****
Let's focus on routes that run at least 150 trips per year. Which ones have the highest percentage of post-bell arrivals and are at least 25% late?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
routes.min.150 <- rawdata %>%
  group_by(Route) %>%
  summarise(arrivals = n()) %>%
  filter(arrivals >= 150)

late.routes <- rawdata %>%
  filter(
    !is.na(bell_late),
    bell_late > 0
  ) %>%
  group_by(Route) %>%
  summarise(
    post.bell.arrivals = n(),
    median.bell_late = median(bell_late)
  )

routes.schools <- rawdata %>%
  group_by(Route) %>%
  summarise(
    school = unique(school)[[1]],
    schtype = unique(schtype)[[1]]
  )

late.routes.pct <- routes.min.150 %>%
  inner_join(late.routes, by=c('Route')) %>%
  mutate(post.bell.pct = 100 * post.bell.arrivals/arrivals) %>%
  inner_join(routes.schools, by=c('Route')) %>%
  ungroup() %>%
  filter(
    schtype == 'BPS',
    post.bell.pct >= 25
    ) %>%
  group_by(school) %>%
  arrange(desc(post.bell.pct))

late.routes.pct %>%
  knitr::kable(caption = 'Year-round routes with highest post-bell arrival percentage')
```

****
So far we have been analysing the data starting at routes. Let's use schools and schtype as starting point. First chart: number of post-bell arrivals by school type.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
late.schools <- rawdata %>%
  filter(
    !is.na(bell_late),
    bell_late > 0,
    school != '',
    schtype != ''
  ) %>%
  group_by(school, schtype) %>%
  summarise(
    post.bell.arrivals = n(),
    median.bell_late = median(bell_late) * 1.0
  )

all.schools <- rawdata %>%
  filter(
    school != '',
    schtype != ''
  ) %>%
  group_by(school, schtype) %>%
  summarise(arrivals = n())

late.schools.pct <- late.schools %>%
  inner_join(all.schools, by=c('school', 'schtype')) %>%
  mutate(post.bell.pct = 100 * post.bell.arrivals / arrivals) %>%
  ungroup() %>%
  arrange(desc(post.bell.pct))

late.schools.pct %>%
  group_by(schtype) %>%
  summarise(post.bell.arrivals = sum(post.bell.arrivals)) %>%
  arrange(desc(post.bell.arrivals)) %>%
  ggplot() +
  geom_bar(aes(schtype, post.bell.arrivals), stat='identity') +
  coord_flip()
```

****
Which BPS schools have the highest percentage of post-bell arrivals?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
late.schools.pct %>%
  filter(schtype == 'BPS') %>%
  select(-schtype) %>%
  head(20) %>%
  knitr::kable(caption = 'BPS schools with highest post-bell arrival percentage')
```

****
Which year-round routes reported no arrival data?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
routes.no.data <- rawdata %>%
  filter(errortype == 'No Arrival Data') %>%
  group_by(Route) %>%
  summarise(no.data.count = n())

routes.min.150 %>%
  inner_join(routes.no.data, by=c('Route')) %>%
  inner_join(routes.schools, by=c('Route')) %>%
  ungroup() %>%
  arrange(desc(no.data.count)) %>%
  head(10) %>%
  knitr::kable(caption='Year-round routes with most "No Arrival Data"')
```

****
Which schools reported no arrival data?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == 'No Arrival Data') %>%
  group_by(school, schtype) %>%
  summarise(no.data.count = n()) %>%
  inner_join(all.schools, by=c('school', 'schtype')) %>%
  mutate(no.data.pct = 100*no.data.count/arrivals) %>%
  ungroup() %>%
  arrange(desc(no.data.count)) %>%
  head(10) %>%
  knitr::kable(caption='Schools with most "No Arrival Data"')
```

****
What days/routes stand out as really bad bus commutes? Where were they going? What day of the week was it?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
routes.min.150 %>%
  inner_join(rawdata, by=c('Route')) %>%
  filter(!is.na(bell_late), bell_late > 0) %>%
  arrange(desc(bell_late)) %>%
  select(Route, bell_late, school, schtype, students, date, dow) %>%
  head(10) %>%
  knitr::kable(caption='Highest post-bell arrival for year-round routes')
```

****
What percentage of these buses reached the school before or at their scheduled time? Before or at the bell?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(!is.na(bell_late)) %>%
  mutate(bell.late = bell_late > 0) %>%
  group_by(bell.late) %>%
  tally() %>%
  mutate(pct = 100*n/sum(n)) %>%
  knitr::kable(caption='Percent of trips arriving after bell')

rawdata %>%
  filter(!is.na(sched_late)) %>%
  mutate(sched.late = sched_late > 0) %>%
  group_by(sched.late) %>%
  tally() %>%
  mutate(pct = 100*n/sum(n)) %>%
  knitr::kable(caption='Percent of trips arriving after scheduled arrival')
```

****
BPS says they use **bell_late** to determine if a bus was late. But this assumes kids need no time to travel from the bus to the classroom (teleportation?). Perhaps a better measure is 10 minutes before school starts. Before we do that, however, let's look at the distribution of the window between **PlannedAnchorTime** and **belltime**.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(errortype == '', !is.na(PlannedAnchorTime), !is.na(belltime), !is.na(ActualArrivalTime)) %>%
  mutate(
    plannedWindow = (hm(PlannedAnchorTime) - hm(belltime))/minutes(1),
    actualWindow = (hm(ActualArrivalTime) - hm(belltime))/minutes(1)
  )

ggplot(data %>% select(plannedWindow, actualWindow) %>% melt(), aes(value)) +
  geom_histogram() +
  scale_y_sqrt(label=comma) +
  facet_grid(variable ~ .) +
  ggtitle('Distribution of planned and actual window (time between bus arrival and bell)') +
  xlab('positive means late, negative means early. (minutes)') +
  ylab('number of buses')
```

****
What is **VersatransDelay**? Let's compare it to **ActualArrivalTime** - **PlannedAnchorTime**.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  select(VersatransDelay, ActualArrivalTime, PlannedAnchorTime, errortype) %>%
  filter(errortype == '', !is.na(VersatransDelay), !is.na(ActualArrivalTime), !is.na(PlannedAnchorTime)) %>%
  mutate(delta = (hm(ActualArrivalTime) - hm(PlannedAnchorTime))/minutes(1)) %>%
  ggplot(aes(VersatransDelay)) +
  geom_histogram() +
  scale_y_sqrt() +
  ggtitle('Distribution of lateness, when late means arriving after planned arrival') +
  xlab('minutes after planned arrival') +
  ylab('number of trips (square root scale)')
```

****
What's going on with PlannedStartTime and ActualArrivalTime - does that mean trip duration?
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '', !is.na(PlannedStartTime), !is.na(ActualArrivalTime)) %>%
  mutate(duration = (hm(ActualArrivalTime) - hm(PlannedStartTime))/minutes(1)) %>%
  ggplot(aes(duration)) +
  geom_histogram() +
  scale_y_sqrt() +
  ggtitle('Distribution of lateness, when late means arriving after planned arrival') +
  ylab('number of trips (square root scale)') +
  xlab('duration (minutes)')
```

****
Correlation between trip duration and number of students.
```{r, results='asis', fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '', !is.na(PlannedStartTime), !is.na(ActualArrivalTime)) %>%
  mutate(duration = (hm(ActualArrivalTime) - hm(PlannedStartTime))/minutes(1)) %>%
  ggplot(aes(duration, students)) +
  stat_binhex()
```

****
Distribution of students per bus.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '') %>%
  ggplot(aes(students)) +
  geom_histogram() +
  scale_y_sqrt() +
  ggtitle('Distribution of students per bus') +
  ylab('number of trips (square root scale)') +
  xlab('students')
```





****
How does the number of student on board play a role, if at all?
```{r, results='asis', fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    !is.na(bell_late),
    !is.na(students)
  ) %>%
  ggplot(aes(students, bell_late)) +
  stat_binhex() +
  ggtitle('Correlation of students to bell_late') +
  xlab('students') +
  ylab('bell_late (minutes)')
```

****
There is no correlation between students and bell_late. Let's look at the number of late students per day.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    !is.na(bell_late),
    bell_late > 0) %>%
  mutate(date = dmy(date)) %>%
  group_by(date, school) %>%
  summarise(students = sum(students)) %>%
  ggplot(aes(date, students)) +
  geom_bar(stat='identity') +
  ggtitle('Post-bell students per day')
```

****
Which schools had the most number of post-bell students?
```{r, results='asis', fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
late.students.per.day <- rawdata %>%
  filter(
    !is.na(bell_late),
    bell_late > 0) %>%
  mutate(date = floor_date(dmy(date), 'month')) %>%
  group_by(date, school) %>%
  summarise(students = sum(students))

top.late.schools <- late.students.per.day %>%
  group_by(school) %>%
  summarise(total = sum(students)) %>%
  ungroup() %>%
  arrange(desc(total)) %>%
  head(30)

top.late.schools %>%
  inner_join(late.students.per.day, by=c('school')) %>%
  ggplot(aes(date, students)) +
  geom_bar(stat='identity') +
  facet_wrap(~ school, ncol=6) +
  ggtitle('Schools with most post-bell students') +
  xlab('month')
```

****
Which schools had the most number of post-bell student-minutes?
```{r, results='asis', fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
student.minutes.per.day <- rawdata %>%
  filter(
    !is.na(students),
    !is.na(bell_late),
    bell_late > 0) %>%
  mutate(student.minutes = students * bell_late) %>%
  mutate(date = floor_date(dmy(date), 'month')) %>%
  group_by(date, school) %>%
  summarise(student.minutes = sum(student.minutes))
  
top.late.schools <- student.minutes.per.day %>%
  group_by(school) %>%
  summarise(total = sum(student.minutes)) %>%
  ungroup() %>%
  arrange(desc(total)) %>%
  head(30)

top.late.schools %>%
  inner_join(student.minutes.per.day, by=c('school')) %>%
  ggplot(aes(date, student.minutes)) +
  geom_bar(stat='identity') +
  facet_wrap(~ school, ncol=6) +
  ggtitle('Schools with most post-bell student.minutes') +
  xlab('month')
```

****
How many buses were late, up to 29 min?
```{r, results='asis', fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    !is.na(bell_late),
    bell_late > 0,
    bell_late < 30) %>%
  tally() 
```

****
How many buses were more than 29 min late?
```{r, results='asis', fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
rawdata %>%
  filter(
    !is.na(bell_late),
    bell_late > 29) %>%
  tally() 
```

****
```{r, results='asis', fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    school != ''
  ) %>%
  mutate(is.bell.late = bell_late > 0) %>%
  arrange(school) %>%
  group_by(school, is.bell.late) %>%
  tally()

all_arrivals <- data %>%
  group_by(school) %>%
  tally() %>%
  rename(all = n)

late_arrivals <- data %>%
  filter(is.bell.late) %>%
  select(-is.bell.late) %>%
  rename(late = n) %>%
  inner_join(all_arrivals, by=c('school')) %>%
  mutate(percent = late/all) %>%
  ungroup() %>%
  arrange(school) 

write.csv(late_arrivals %>% select(school), 'late_arrivals.csv', row.names = FALSE)
```

