---
title: "schoolbuses - main"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
```

****
This is a story about buses. Boston Public School buses.

On any given day, hundreds of buses carry thousands of kids to school. Here's September 4, the first day of classes. [Animate bar height and update bus trip count]  
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '') %>%
  mutate(date = dmy(date)) %>%
  group_by(date) %>%
  tally() %>%
  mutate(is.first = row_number() == 1) %>%
  ungroup() %>%
  head(10) %>%
  arrange(date) %>%
  ggplot(aes(date, n, alpha=is.first)) +
  geom_bar(stat='identity') +
  theme(legend.position='none') +
  scale_alpha_manual(values=c(0, 1)) +
  ylab('bus trips')
```

****
And here's the rest of the year.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '') %>%
  mutate(date = dmy(date)) %>%
  group_by(date) %>%
  tally() %>%
  mutate(is.first = row_number() == 1) %>%
  ungroup() %>%
  arrange(date) %>%
  ggplot(aes(date, n)) +
  geom_bar(stat='identity') +
  theme(legend.position='none') +
  scale_alpha_manual(values=c(0, 1)) +
  ylab('bus trips')
```

****
Buses often arrive late. On the first day of school, 524 buses showed up after the bell. [Animate dots alpha and y, and text next to dots]
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(
    PlannedAnchorTime <= belltime,
    month(date) == 9,
    day(date) < 10
  ) %>%
  mutate(
    is.first = day(date) == 4,
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late, alpha=is.first)) +
  geom_point(size=1) +
  scale_colour_manual(values=c('grey80', '#ea212d')) +
  scale_alpha_manual(values=c(0, 1)) +
  theme(legend.position='none')
```

****
Over the school year, as drivers become familiar with their routes, tardiness decreases. But buses continue to arrive late, often by over 30 minutes.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(
    PlannedAnchorTime <= belltime
  ) %>%
  mutate(
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late)) +
  geom_point(size=0.5) +
  scale_colour_manual(values=c('grey80', '#ea212d')) +
  theme(legend.position='none')
```

****
For some schools, late buses are a constant. Take Quincy Elementary. Over 18% of the buses arrived after the bell.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime),
    school == 'Quincy Elementary'
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(PlannedAnchorTime <= belltime) %>%
  mutate(
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late)) +
  geom_point(size=1) +
  scale_colour_manual(values=c('grey80', '#ea212d')) +
  theme(legend.position='none')
```