```{r setup, cache=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE)
```

---
title: "schoolbuses - main"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
```

### Data analysis

****
Why is BPS using **bell_late** to determine tardiness? Why not 10 minutes before school starts? That would seem to give kids some time to get off the bus and on to their first class. Let's create a new variable, **ten_late**, and calculate overall tardiness by three definitions of late:

- **bell late**: did the bus make the school bell?
- **ten late**: did the bus arrive ten minutes before the school bell?
- **scheduled late**: did the bus arrive past its scheduled (by the software) date?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(sched_late),
    !is.na(belltime),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    date = floor_date(dmy(date), 'day'),
    ten_late = ((hm(ActualArrivalTime) - hm(belltime))/minutes(1) + 10),
    is.bell.late = bell_late > 0,
    is.ten.late = ten_late > 0,
    is.sched.late = sched_late > 0
  )

data.days %>%
  group_by(is.bell.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.ten.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.sched.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()
```

****
Let's plot weekly tardiness by these three variables.
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
bell.late <- data.days %>%
  group_by(date, is.bell.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(bell.late = n/sum(n)) %>%
  filter(is.bell.late) %>%
  select(-is.bell.late, -n)

ten.late <- data.days %>%
  group_by(date, is.ten.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(ten.late = n/sum(n)) %>%
  filter(is.ten.late) %>%
  select(-is.ten.late, -n)

sched.late <- data.days %>%
  group_by(date, is.sched.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(sched.late = n/sum(n)) %>%
  filter(is.sched.late) %>%
  select(-is.sched.late, -n)

late <- bell.late %>%
  inner_join(ten.late, by=c('date')) %>%
  inner_join(sched.late, by=c('date')) %>%
  melt(c('date'))

ggplot(late, aes(date, value, fill=variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('Percent of tardy buses by various definitions of "late"') +
  scale_y_continuous(labels=percent) +
  theme(axis.title.y = element_blank()) +
  xlab('day')
```

****
Let's focus on the worst schools.
{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
timeHMS_formatter <- function(x) {
    h <- floor(x/60)
    m <- floor(x %% 60)
    s <- round(60*(x %% 1))                   # Round to nearest second
    lab <- sprintf('%02d:%02d', h, m) # Format the strings as HH:MM:SS
    lab <- gsub('^00:', '', lab)              # Remove leading 00: if present
    lab <- gsub('^0', '', lab)                # Remove leading 0 if present
}

worst_schools <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    bell_late > 0,
    schtype == 'BPS'
  ) %>%
  group_by(school) %>%
  tally() %>%
  arrange(desc(n)) %>%
  head(1)

worst_schools

data <- worst_schools %>%
  rename(late.arrivals = n) %>%
  select(school, late.arrivals) %>%
  inner_join(rawdata, by=c('school')) %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    is.late = bell_late > 0,
    date = dmy(date),
    ActualArrivalTime = hm(ActualArrivalTime)/hours(1)
  )

data$school <- factor(data$school, levels = data$school[order(-data$late.arrivals)])

ggplot(data, aes(date, ActualArrivalTime, color=is.late, size=is.late)) +
  geom_point() +
  facet_grid(. ~ school) +
  scale_colour_manual(values=c('grey70', '#ea212d')) +
  scale_size_manual(values=c(0.75, 0.9)) +
  theme(legend.position='none') +
  ylab('arrival') +
#  scale_y_continuous(label=timeHMS_formatter) +
  ggtitle('Late arrivals (in red) at worst schools')
```

****
In 2013-14, over 18,000 bus trips arrived after the bell.
{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(PlannedAnchorTime <= belltime) %>%
  mutate(
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late, size=is.late)) +
  geom_point() +
  scale_colour_manual(values=c('grey70', '#ea212d')) +
  scale_size_manual(values=c(0.5, 0.8)) +
  theme(legend.position='none')
```

****
For some schools, late buses are a constant. Take Quincy Elementary. Over 18% of the buses arrived after the bell.
{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime),
    school == 'Quincy Elementary'
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(PlannedAnchorTime <= belltime) %>%
  mutate(
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late, size=is.late)) +
  geom_point() +
  scale_colour_manual(values=c('grey70', '#ea212d')) +
  scale_size_manual(values=c(0.7, 1)) +
  theme(legend.position='none')
```





****
This is a story about buses. Boston Public School buses.

On any given day, hundreds of buses carry thousands of kids to school. Here's September 4, the first day of classes. [Animate bar height and update bus trip count]  
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '') %>%
  mutate(date = dmy(date)) %>%
  group_by(date) %>%
  tally() %>%
  mutate(is.first = row_number() == 1) %>%
  ungroup() %>%
  head(10) %>%
  arrange(date) %>%
  ggplot(aes(date, n, alpha=is.first)) +
  geom_bar(stat='identity') +
  theme(legend.position='none') +
  scale_alpha_manual(values=c(0, 1)) +
  ylab('bus trips')
```

****
And here's the rest of the year.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
rawdata %>%
  filter(errortype == '') %>%
  mutate(date = dmy(date)) %>%
  group_by(date) %>%
  tally() %>%
  mutate(is.first = row_number() == 1) %>%
  ungroup() %>%
  arrange(date) %>%
  ggplot(aes(date, n)) +
  geom_bar(stat='identity') +
  theme(legend.position='none') +
  scale_alpha_manual(values=c(0, 1)) +
  ylab('bus trips')
```

****
Buses often arrive late. On the first day of school, 524 buses showed up after the bell. [Animate dots alpha and y, and text next to dots]
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(
    PlannedAnchorTime <= belltime,
    month(date) == 9,
    day(date) < 10
  ) %>%
  mutate(
    is.first = day(date) == 4,
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late, alpha=is.first)) +
  geom_point(size=1) +
  scale_colour_manual(values=c('grey70', '#ea212d')) +
  scale_alpha_manual(values=c(0, 1)) +
  theme(legend.position='none')
```

****
Over the school year, as drivers become familiar with their routes, tardiness decreases. But buses continue to arrive late, often by over 30 minutes.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(
    PlannedAnchorTime <= belltime
  ) %>%
  mutate(
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime
  )

ggplot(data, aes(date, late.minutes, color=is.late)) +
  geom_point(size=1) +
  scale_colour_manual(values=c('grey70', '#ea212d')) +
  theme(legend.position='none')
```

****
often by over 30 minutes.
```{r, results='asis', fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
data <- rawdata %>%
  filter(
    errortype == '',
    !is.na(PlannedAnchorTime),
    !is.na(ActualArrivalTime),
    !is.na(belltime)
  ) %>%
  mutate(
    date = dmy(date),
    PlannedAnchorTime = hm(PlannedAnchorTime)/minutes(1),
    ActualArrivalTime = hm(ActualArrivalTime)/minutes(1),
    belltime = hm(belltime)/minutes(1)
  ) %>%
  filter(
    PlannedAnchorTime <= belltime
  ) %>%
  mutate(
    late.minutes = ActualArrivalTime - belltime,
    is.late = ActualArrivalTime > belltime,
    really.late = late.minutes > 29
  ) %>%
  filter(really.late)

ggplot(data, aes(date, late.minutes, color=really.late)) +
  geom_point(size=1) +
  scale_colour_manual(values=c('#ea212d')) +
  theme(legend.position='none')
```







