```{r setup, cache=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(echo=FALSE)
```
---
title: "schoolbuses - main"
author: "Gabriel Florit"
output:
  html_document:
    self_contained: false
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(lubridate)
setwd("~/Documents/dev/schoolbuses/data")
```

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=TRUE}
rawdata <- read.table('input/SY1314_AM_Bus_Data_for_the_Globe.txt', sep='\t', header=TRUE)
```

### Summary

In the 2013-14 school year, Boston Public School buses arrived after the bell 7% of the time. However, they arrived with less than 10 minutes before the bell, 29% of the time. What impact does this have on kids?

- If they eat breakfast before bell, then only arriving at bell is not enough.
- If they eat breakfast after bell, then it seems schools are accepting buses will be tardy, and are incorporating chronic bus lateness into the school structure. 
- According to BPS media relations director Denise Snyder, schools offer a "Grab and go" alternative for students that missed breakfast. What is in this "Grab and go" and how does it differ from regular breakfast?

### Data analysis

****
Why is BPS using **bell_late** to determine tardiness? Why not 10 minutes before school starts? That would seem to give kids some time to get off the bus and on to their first class. Let's create a new variable, **ten_late**, and calculate overall tardiness by three definitions of late:

- **bell late**: did the bus make the school bell?
- **ten late**: did the bus arrive ten minutes before the school bell?
- **scheduled late**: did the bus arrive past its scheduled (by the software) date?
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
data.days <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(sched_late),
    !is.na(belltime),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    date = floor_date(dmy(date), 'day'),
    ten_late = ((hm(ActualArrivalTime) - hm(belltime))/minutes(1) + 10),
    is.bell.late = bell_late > 0,
    is.ten.late = ten_late > 0,
    is.sched.late = sched_late > 0
  )

data.days %>%
  group_by(is.bell.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.ten.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()

data.days %>%
  group_by(is.sched.late) %>%
  tally() %>%
  mutate(percent = n/sum(n)) %>%
  knitr::kable()
```

****
Let's plot weekly tardiness by these three variables.
```{r, results='asis', fig.width=10, fig.height=3, warning=FALSE, message=FALSE}
bell.late <- data.days %>%
  group_by(date, is.bell.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(bell.late = n/sum(n)) %>%
  filter(is.bell.late) %>%
  select(-is.bell.late, -n)

ten.late <- data.days %>%
  group_by(date, is.ten.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(ten.late = n/sum(n)) %>%
  filter(is.ten.late) %>%
  select(-is.ten.late, -n)

sched.late <- data.days %>%
  group_by(date, is.sched.late) %>%
  tally() %>%
  group_by(date) %>%
  mutate(sched.late = n/sum(n)) %>%
  filter(is.sched.late) %>%
  select(-is.sched.late, -n)

late <- bell.late %>%
  inner_join(ten.late, by=c('date')) %>%
  inner_join(sched.late, by=c('date')) %>%
  melt(c('date'))

ggplot(late, aes(date, value, fill=variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('Percent of tardy buses by various definitions of "late"') +
  scale_y_continuous(labels=percent) +
  theme(axis.title.y = element_blank()) +
  xlab('day')
```

****
Let's focus on the worst schools.
```{r, results='asis', fig.width=10, fig.height=20, warning=FALSE, message=FALSE}
timeHMS_formatter <- function(x) {
    h <- floor(x/60)
    m <- floor(x %% 60)
    s <- round(60*(x %% 1))                   # Round to nearest second
    lab <- sprintf('%02d:%02d', h, m) # Format the strings as HH:MM:SS
    lab <- gsub('^00:', '', lab)              # Remove leading 00: if present
    lab <- gsub('^0', '', lab)                # Remove leading 0 if present
}

worst_schools <- rawdata %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    bell_late > 0
  ) %>%
  group_by(school) %>%
  tally() %>%
  arrange(desc(n)) %>%
  head(10)

worst_schools %>%
  select(school) %>%
  inner_join(rawdata, by=c('school')) %>%
  filter(
    errortype == '',
    !is.na(bell_late),
    !is.na(ActualArrivalTime)
  ) %>%
  mutate(
    is.late = bell_late > 0,
    date = dmy(date),
    ActualArrivalTime = hm(ActualArrivalTime)/hours(1)
  ) %>%
  ggplot(aes(date, ActualArrivalTime, color=is.late, size=is.late)) +
  geom_point() +
  facet_grid(school ~ ., scales='free_y') +
  scale_colour_manual(values=c('grey70', '#ea212d')) +
  scale_size_manual(values=c(0.75, 0.9)) +
  theme(legend.position='none') +
  ylab('arrival') +
#  scale_y_continuous(label=timeHMS_formatter) +
  ggtitle('Late arrivals (in red) at worst schools')
```
